---
title: "Petitions in ggplot and Observable Plot"
subtitle: "Practising the dark arts of Observable JS in Quarto with a simple bar chart"
author: "Sharon Howard"
date: "2025-07-03"
categories: [interactive, methods, petitioning]
execute: 
  warning: false
  message: false
---

## Introduction

I've started exploring [Quarto's built in support for Observable JS](https://quarto.org/docs/interactive/ojs/) to make interactive charts. Here I'm working through converting [a bar chart of petitioning topics I made earlier](https://the-power-of-petitioning.github.io/topics.html#overview-of-topics) in ggplot to use [Observable Plot](https://observablehq.com/plot/).


```{r}
library(tidyverse)

# load data package with the cheshire_petitions data
library(mindseyedata)

# make a custom colour palette
topic_colours <- c("alehouse" = "#a6cee3",	"charitable brief" = "#1f78b4",	
                   "cottage" =	"#b2df8a", "dissenting worship" =	"#33a02c", 
                   "employment" =	"#ff7f00", "imprisoned debtors" =	"#fdbf6f", 
                   "litigation" = "#e31a1c", "military relief" = "#fb9a99", 
                   "officeholding" =	"#ffed6f", "other" = "#8c510a", 
                   "paternity"	=	"#cab2d6", "poor relief" = "#6a3d9a", 
                   "rates" = "#80cdc1")

# slightly adjust topics in the petitions
# (there's only a single instance of imprisoned debtors in Cheshire)
cheshire_petitions <-
cheshire_petitions |>
  mutate(topic = if_else(topic=="imprisoned debtors", "other", topic)) |>
  select(petition_id, topic, petition_type, response)


#cheshire_petitions
```


## Petitioning topics in ggplot

This looks slightly different from the original as I'm only using Cheshire petitions, not the full TPOP set. But the ggplot code is almost identical.


```{r}
cheshire_petitions |>
  count(topic) |>
  mutate(topic=fct_reorder(topic, n)) |>
  ggplot(aes(x=n, y=topic, fill=topic)) +
  geom_col() +
  scale_fill_manual(values = topic_colours)  +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(y=NULL, x="number of petitions", 
       title="Popularity of topics in Cheshire QS petitions")
```

## OJS

Here is a neat thing about OJS, compared to R: you don't have to define data before you use it. Cells (`ojs` rather than `r` chunks) can go in any order. 

Here's the petitions data converted to OJS format, though the actual code to make it is a bit further down this post. R would be most unhappy about this state of affairs. 


```{ojs}
ojsPetitions
```

There are various ways to get data in [formats OJS can use](https://quarto.org/docs/interactive/ojs/data-sources.html). To convert the `cheshire_petitions` R data frame needs two steps:

- `ojs_define()` inside an R chunk to pass the data to Observable
- `transpose()` inside an OJS chunk to switch it to the right format for Plot

```{r}
ojs_define(ojs_petitions = cheshire_petitions)
```

```{ojs}
ojsPetitions = transpose(ojs_petitions)
```

(I tend to name R objects using snake_case and OJS objects with camelCase, so I'm slightly less likely to get confused when working with both kinds of data in the same document.)


## Making the chart
	
The key features of the bar chart that I want to reproduce in Observable Plot:

- horizontal bars
- ordered by frequency rather than alphabetically
- coloured by topic (using a manually defined palette)

(I also added tooltips, just because I can.)

```{ojs}
Plot.plot({
  marginLeft: 100, 
  title: "Popularity of topics in Cheshire QS petitions", 
  x: {grid: true, label: "number of petitions"},
  y: {grid:true, label: null},
  color: topicColours,
  marks: [
    Plot.barX(ojsPetitions, 
                Plot.groupY({x: "count"}, 
                            {y: "topic", 
                              fill: "topic", 
                              tip: true,
                              sort: {y: "x", reverse: true}
                            })
              ),
    Plot.ruleX([0])
  ]
})
```

### notable differences

Both ggplot and Plot often have multiple ways of doing the same thing and I know much less about all the options in Plot, so some differences might be more apparent than real.

For the ggplot, I aggregated the data (dplyr::count()) and reordered it (forcats::fct_reorder()) *before* starting the ggplot code. I could certainly do the aggregation for Plot beforehand too but I'd need to look up how, and I have absolutely no idea if there's an equivalent of fct_reorder. It seems easier to do both of those inside the Plot (groupY() and sort).

When using a colour scale, gpplot makes a legend by default and has to be told not to, whereas you have to tell Plot to make one.

The default aspect ratio is clearly very different; I'd need to check how to change that in Plot (and also how to change the title text size). Also, ggplot automatically makes space for long labels but Plot doesn't, so you have to set a margin.

Finally, I needed to reformat the topic_colours list for the colour palette, into (to me, anyway) a less transparent format. I'll look into how to wrangle the list properly for Plot.scale(); for now I rewrote it by hand.

```{ojs}
topicColours = Plot.scale({
		color: {
			range: ["#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#ff7f00", "#fdbf6f", "#e31a1c", "#fb9a99", "#ffed6f", "#8c510a",  "#cab2d6",  "#6a3d9a",  "#80cdc1"], 
			domain: ["alehouse", "charitable brief", "cottage", "dissenting worship", "employment", "imprisoned debtors", "litigation", "military relief", "officeholding", "other",  "paternity", "poor relief", "rates"]
		}
	});
```


## Resources

- [Quarto Guide on Observable JS](https://quarto.org/docs/interactive/ojs/)
- [Observable for R Users](https://nrennie.rbind.io/blog/observable-r-users/)
- I found this [Observable Notebook on converting ggplot to Plot](https://observablehq.com/@mayagans/from-ggplot2-to-plot) particularly helpful
- specific Observable Plot docs
  - [bar mark](https://observablehq.com/plot/marks/bar)
  - [group transform](https://observablehq.com/plot/transforms/group)

