---
title: 'Old Bailey Proceedings Part 1: Offences'
author: Sharon Howard
date: '2018-04-17'
slug: old-bailey-proceedings-part-1-offences
categories: 
  - OBO
  - crime
ihmefilename: "201804_obo_1"
execute: 
  warning: false
  message: false
  echo: false
---

```{r}
#| include: false
# review July 2026
# tidied old clunky code
```



Introduction
-------


What are the *Old Bailey Proceedings*? That's the name most commonly given to a series of trial reports that were published from 1674-1913:

>the largest body of texts detailing the lives of non-elite people ever published, containing 197,745 criminal trials held at London's central criminal court.

They have been digitised by the [Old Bailey Online project (OBO)](https://www.oldbaileyonline.org). 


I've been project manager for this and a number of spin-off projects (which I'll undoubtedly write about in future posts) since 2006. And yet I only started to really dig into the Proceedings data quite recently. This is because it consists of more than 2000 intimidatingly complicated XML files, reflecting the complexity of a criminal trial - there can be multiple defendants, multiple charged offences and multiple outcomes. The [central aim of the project from its conception](https://earlymodernnotes.wordpress.com/2013/04/13/tales-of-the-unexpected-or-what-can-happen-when-you-let-a-bunch-of-criminals-loose-on-the-internet/) was to accurately represent this complexity as well as provide searchable full text. 

In fact, I spent several years cheerfully encouraging others to use our data while I had no real idea how to go about doing so myself. In 2011, the project released the Old Bailey API, and I started to [tinker with that](http://sharonhoward.org/ob/), but I didn't really get very far, until a couple of years ago I finally bought myself a book on XQuery and got down to it. And then I started to discover *exactly* how complicated the XML is. (So I've also been thinking about ways to make it more accessible; putting the XML files on our institutional repository is a good start but it's really just a start.)


The data
----

There are two facets to the the Proceedings data: firstly, structured markup that enables searching and quantitative analysis of many aspects of the published trials (especially the characteristics of defendants, offences, jury verdicts, sentences); and secondly, the full text of the reports, amounting to more than 125 million words in total.

My first two posts are tasters of a few of the structured data categories: here, I'll look at how offences tried at the Old Bailey changed over the 250 years documented in the Proceedings; in the second post, I'll look at defendants' gender and offending. In subsequent posts I'll start to explore the text of trial reports.

The variables I'll be working with:

* year of trial
* offence count - number of offences charged per trial
* offence categories (first listed if there were multiple offences)
* defendant count - number of defendants per trial. This has also been used to create:
* defendant type - single or group (this is really just for convenience)

Along the way, we'll look at several different ways of visualising data, and their different vices and virtues. The most familiar to historians will be text tables, bar charts and line graphs. But I'm also going to look at something most historians are less likely to have used: heatmaps.

(There's more detail in the R code and at the end of the post about the dataset.) 


```{r}
#load r packages 

library(scales)  # additional scaling functions mainly for ggplot (eg %-ages)
library(knitr)  # for kable (nicer tables)
library(kableExtra)  

library(tidyverse)  # the core Tidy toolkit

library(ggthemes)
theme_set(theme_minimal()) # set preferred ggplot theme 

# get data

# NB this is OBO v7.2 data and is now slightly out of date, but changes in the update to v8.0 (March 2018) were fairly minor

# read the data in with read_csv (instead of read.csv)
# nb: using na="NULL", for data imported from MySQL this converts NULL to NA

obo <- read_csv(here::here("site_data/obo/obo_trials_20180402.csv"), na="NULL" )



# notes on the data 
# (see also the main text intro and appendix)

# *id*: unique ID in dataset
# *tid*: OBO trial reference; if you want to look up a particular trial in a browser, just append the reference to the base URL https://www.oldbaileyonline.org/record/
# *sess_date*: OBO session date reference (yyyymmdd)
# *trial_sort*: the number of the trial in the session
# *defcount*: number of defendants in the trial
# *offcatcount*: number of offences charged in the trial
# *genders*: defendant(s) gender(s), list separator ||
# *offCats*: category/ies of offence(s) charged, list separator ||
# *offcat1*: offence category if only one offence, first offence listed if multiple
# *gender1*: same thing for defendant gender
# *wordcount*: trial report word count (I think this really needs re-doing)



# data preparation 

# add columns: yyyy-mm-dd date column; single/group defendant type (this is really just for convenience)
obo_prep <- obo |>
  mutate( year_month_day = parse_date(as.character(sess_date), format = "%Y%m%d"),
          def_type =  ifelse(defcount>1, "group", "single")
          )

# filter out handful of trials without any offence category
obo_prep <- obo_prep |>
  filter(offcat1 != '') |>
  droplevels() # this completely removes filtered "levels", which matters for the mosaic plots; might as well do it from the start


# new dataframe obo_trials, with just the variables I'll actually need in this post
# and filter out 1674 and 1913 because they're incomplete years
obo_trials <- obo_prep |> select(id, year_month_day, offcatcount, offcat1, defcount, def_type) 

```


Here's a little slice of what it looks like:
 

```{r}
kable(obo_trials |> slice(197500:197515) )
```


First views
----

Having said that trials can be varied in terms of numbers of defendants, offences and so on, a histogram makes it clear that in fact the vast majority have a single defendant accused of a single offence. The highest number of defendants in one trial is actually [``r max(obo_trials$defcount, na.rm = TRUE)``](https://www.oldbaileyonline.org/record/t18210411-15), but there are so few trials with more than 8 defendants that the bars aren't even visible on the chart; a similar thing happens with offence counts. 

But I'm not *entirely* going to forget that there are ``r obo_trials |> filter(defcount>1) |> count() `` trials (``r paste0(round(obo_trials |> filter(def_type == "group") |> tally() / tally(obo_trials) * 100, 1), "%") `` of the total) that do have more than one defendant - they might not have the same characteristics as single defendant trials.



```{r obo-1-histogram-defendants-1}
# histogram of defendants per trial, filtering out NAs
  obo_trials |> 
  filter(!is.na(defcount)) |>
  ggplot(aes(defcount) ) +
  geom_histogram(bins=30) +
  scale_x_continuous(breaks = c(1,3,5,10,15,20,25,30, 35)) +
  labs(title="Number of defendants per trial", x="number of defendants", y="count of trials")

```



```{r obo-1-histogram-offences-1}

#histogram of offences charged per trial, filtering out NAs
  obo_trials |> 
  filter(!is.na(offcatcount))   |>
  ggplot(aes(offcatcount) ) +
  geom_histogram(bins=15) +
  scale_x_continuous(breaks = c(1,3,5,10,20)) +
  labs(title="Number of charged offences per trial", x="number of offences", y="count of trials")

```



There are nine broad offence categories in the OBO data (with several sub-categories in each category which I won't consider today):


```{r}
# showing summarised data in a table (use dplyr count())
kable(obo_trials |> count(offcat1))
```

A quick explanation of the categories ([more info](https://www.oldbaileyonline.org/about/crimes)):

1. **breakingPeace**: various sorts of disorderly behaviour including assault and riot
2. **damage**: damage to property, eg arson, smashing windows
3. **deception**: mainly fraud and forgery, also perjury
4. **kill**: murder, manslaughter, infanticide, etc
5. **miscellaneous**: anything that doesn't fit into the main categories or couldn't be categorised (because it's so assorted I'll be pretty much ignoring it)
6. **royalOffences**: crimes against the authority of the Crown, government or Church, eg treason, sedition and coining
7. **sexual**: includes rape, sexual assault, sodomy, bigamy, keeping a brothel
8. **theft**: non-violent theft including larceny, burglary and housebreaking, pocket picking, shoplifting, animal theft, as well as receiving stolen goods
9. **violentTheft**: theft with violence - robbery and highway robbery

As you can see, theft dominates completely - it accounts for ``r 
paste0(round(obo_trials |> filter(offcat1 == "theft") |> tally() / tally(obo_trials) * 100 , 1 ), '%') `` of all trials. This often causes major headaches for visualisation - theft can just overwhelm everything. 

Before I move on to that problem, though, I want to go back to the question of possible differences between single- and multiple-defendant trials, and I can start by looking at whether they're tried for different types of offence. The answer is: not that much variation in most categories. Robbery, however, tends to be a group crime (put away your romantic notions of lone gentlemen highwaymen), and sexual offences tend to be committed alone. 

```{r obo-1-bar-trials-1}
# this is very strange code. why facet? cba to 
  obo_trials |> 
  filter(!is.na(def_type)) |> 
  count(def_type, offcat1) |>
ggplot(aes(x="", y=n, fill=def_type) ) +
  geom_bar(stat = "identity", position = "fill") +
  facet_grid(~offcat1) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_tableau() +
  labs(x="", y="")
```



Offences tried over time
-----------

Let's start with an overview of annual numbers of trials from 1674 to 1913. There are quite a lot of gaps in the records until about 1715, and the very pronounced dip at the beginning of the 18th century needs to be viewed very cautiously. In addition, in the early decades the Proceedings were not intended to be comprehensive. It's thought that most trials were reported (if only briefly) during the 18th century, though the scale of omissions is not really known. By the beginning of the 19th century, the Proceedings were a semi-official record of the court and gaps should be few and far between.

The most obvious thing in the graph is the *massive* increase in trials from the early 19th century, which very suddenly drops away again after about 1845. Did Londoners become much more criminal for 3-4 decades, and then suddenly stop breaking the law again? Very unlikely! 

The increase can probably be largely accounted for by rapid growth in London's population (I should dig out some population numbers) and quite possibly from expansion of the city's police forces. Explaining the subsequent decrease is simpler: it was the result of changes in policy which channelled the prosecution of less serious crimes away from the Old Bailey (to magistrates' courts). As we'll see in a moment, that had a profound impact not just on numbers but also on the characteristics of prosecuted trials.


```{r obo-1-trials-line-annual-1}
# line graph of annual numbers of trials
obo_trials |>
  select(year_month_day) |>
  mutate(year = year(year_month_day)) |>
  add_count(year) |>
  ggplot(aes(x=year, y=n) ) +
  geom_line() +
  scale_x_continuous(breaks=seq(1650, 1925, by=25),limits=c(1674,1913) ) +
  theme_bw() +
  labs(title="Trials in the Old Bailey, 1674-1913", y="number of trials")

```


I want to look in more detail at the patterns of offending by category within this, but this kind of graph is not ideally suited to the task, because of the excess of theft trials. It's pretty hard to see what's happening with *any* of the other categories in much of this chart:


```{r obo-1-line-year-offcat-1}

# summarise year x offcat
obo_trials_off_year <- 
  obo_trials |>
  mutate(year = year(year_month_day)) |>
  count(year, offcat1, name="num")

# line graph of annual numbers of trials by offence categories
ggplot(obo_trials_off_year 
       , aes(x=year, y=num, colour=offcat1) ) +
  geom_line() +
  theme_bw() +
  labs(title="Trials by offence categories, 1674-1913", y="number of trials in category")

```

The one thing you can see clearly is that theft accounts for most of the mid-19th-century drop in numbers - indeed, the other categories show increases. But it's still hard to pick out any meaningful details. If I repeat the graph but take out theft completely, it's easier to grasp the scale of growth in the other categories, but it's still crowded and hard to pick out individual lines. 

```{r obo-1-trials-annual-line-1}

# line graph of annual numbers of trials by offence categories excl theft
obo_trials_off_year |> filter(offcat1 != "theft") |>
  ggplot(aes(x=year, y=num, colour=offcat1) ) +
  geom_line() +
  theme_bw() +
  labs(title="Trials by offence categories excluding theft, 1674-1913", y="number of trials in category")


```

This is another case where faceting is useful. It neatly separates out all the confusion of the competing lines and highlights the different trends. 

```{r obo-1-trials-offences-annual-line-1}

# line graph of annual numbers of trials by offence categories
obo_trials_off_year |> filter(offcat1 != "theft") |>
ggplot( aes(x=year, y=num, colour=offcat1) ) +
  geom_line() +
  theme_bw() +
  facet_wrap(~offcat1) +
  theme(legend.position="none") +
  labs(title="Trials by offence categories excluding theft, 1674-1913", y="number of trials in category")

```



But still, I'd like a better way to visualise and compare all the categories at once. A heatmap is much better at this than line graphs: each category in each year is represented by a single tile and the darker a tile, the more trials there were in that category. 
 

```{r obo-1-heatmap-offences-year-1, out.width="100%"}
#heatmap of the same data using geom_raster()
  obo_trials_off_year |>
  ggplot(aes(year, offcat1, fill=num)) +
  geom_raster() +
  scale_x_continuous(breaks=seq(1670, 1915,by=10),limits=c(1674,1913), expand=c(0,0) ) +
  scale_y_discrete(expand=c(0,0)) +
  scale_fill_gradient(trans="log10", low = "#fbf9d3", high = "#832525") +
  theme(axis.text.x=element_text(angle=45,hjust=1.0,vjust=1.0), axis.text.y  = element_text(size=8), axis.title.y=element_blank(), panel.background = element_rect(fill = "white", colour = "#fbf9d3"), legend.position = "bottom" ) +
  labs(y="", fill="number of trials", title="Heatmap of OBO trial offence categories 1674-1913") 

```


You need to look quite carefully at the legend for this first example. It uses a [logarithmic scale](https://en.wikipedia.org/wiki/Logarithmic_scale) rather than linear scale. This is particularly useful for data like this which has a very wide range of values. It de-emphasises the rise and fall in theft prosecutions c.1800-1850. But you can now see changes in the smaller offence categories much more clearly. 

One of the most striking short-term blips is in the breakingPeace category in 1780, which is largely down to ``r obo_trials |> filter(year_month_day =="1780-06-28", offcat1 =="breakingPeace") |> count(offcat1) |> select(n) `` trials in the Old Bailey session of [28 June](https://www.oldbaileyonline.org/search/keyword?month_gte=6&month_lte=6&offence=breakingPeace&year_gte=1780&year_lte=1780#results). 18th-century historians may guess that this has something to do with the [Gordon Riots](https://en.wikipedia.org/wiki/Gordon_Riots) in early June 1780. The Gordon Riots were among the most serious and destructive domestic disturbances ever to take place in London; just how much so is highlighted here remarkably well.

The graph suggests other potentially interesting variations. Trials in the deception category seem to form a consistently higher profile from around 1840. The royalOffences category (which largely consists of coining rather than more exciting high treason trials) appears busy in the late 17th century, picks up again briefly in the 1810s and then has another run from c.1850-85. And sexual offences seem to become more numerous from around 1885.

But so far it's all been dealing with numbers, and as we could see in the line graph, those changed a lot - short term and long term - during the period covered by the graph. So I've re-done the heatmap so that each tile now represents the category as a *percentage* of all trials in that year.

```{r obo-1-heatmap-offences-year-2, out.width="100%"}
#category as % of total trials in a year
obo_trials |>
  mutate(year = year(year_month_day)) |>
  add_count(year, name = "n_yr") |>
  count(year, offcat1, n_yr, name="n_yr_offcat") |>
  mutate(pc_yr_offcat = n_yr_offcat/n_yr) |>
  ggplot(aes(year, offcat1, fill=pc_yr_offcat)) +
  geom_raster() +
  scale_x_continuous(breaks=seq(1670, 1915, by=10),limits=c(1674,1913), expand=c(0,0) ) +
  scale_y_discrete(expand=c(0,0)) +
  scale_fill_gradient(low = "#fbf9d3", high = "#832525", labels=percent_format() ) +
  theme(axis.text.x=element_text(angle=45,hjust=1.0,vjust=1.0), axis.text.y  = element_text(size=8), axis.title.y=element_blank(), panel.background = element_rect(fill = "white", colour = "#fbf9d3"), legend.position = "bottom" ) +
  labs(y="", fill="% of trials in year", title="Heatmap of OBO trial offence categories 1674-1913") 

```


This approach reinstates theft pre-eminence and de-emphasises short-term and smaller changes in other categories. Once again theft's growing dominance looks a much more gradual long-term shift than in the first line graph - and again, the suddenness and degree of the mid-19th-century fall is confirmed again. 

It's inevitable that a big decrease in the proportion of trials in one category is going mean other categories must increase, but it's much more noticeable in some categories than others (breakingPeace, deception, royalOffences, sexual), and with varying timings. 

It will come as no great surprise to historians of violence that the killing category declines markedly as a proportion of trials after the end of the 17th century. But even that isn't entirely straightforward. I think it looks more obvious than it did on the first heatmap because of the selective and rather sensationalist reporting of early Proceedings in which murders were more likely to be reported than less 'exciting' crimes.

And beware! In 1706 there is a very dark tile for sexual offences. That's simply because there's only one surviving file for that year, containing a single trial for bigamy. In history data, if something looks wacky, it's usually because there's something up with the data, not the history, and only occasionally because of something really interesting like the Gordon Riots. 


Here's one more perspective using a heatmap. In this one, tiles represent a category in one year as a percentage of the *total for that category*. 

```{r obo-1-heatmap-offences-category-3, out.width="100%"}

#category in year as % of total trials in that category
#heatmap of the same data using geom_raster()
  obo_trials |>
  mutate(year = year(year_month_day)) |>
  add_count(year, name = "n_yr") |>
  add_count(offcat1, name="n_offcat") |>
  count(year, offcat1, n_yr, n_offcat, name="n_yr_offcat") |>  
  mutate(pc_yr_offcat = n_yr_offcat/n_offcat*100) |>
ggplot(aes(year,offcat1, fill=pc_yr_offcat)) +
  geom_raster() +
  scale_x_continuous(breaks=seq(1670, 1915,by=10),limits=c(1674,1913), expand=c(0,0) ) +
  scale_y_discrete(expand=c(0,0)) +
  scale_fill_gradient(low = "#fbf9d3", high = "#832525") +
  #theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust=1.0,vjust=1.0), axis.text.y  = element_text(size=8), axis.title.y=element_blank(), panel.background = element_rect(fill = "white", colour = "#fbf9d3"), legend.position = "bottom" ) +
  labs(y="", fill="% of trials in offence category", title="Heatmap of OBO trial offence categories 1674-1913") 
```


Of course, this view completely flattens out the very different numbers in each category, but it has the great benefit of clarifying patterns *within* categories. We can see the variations in post-1850 trends, and maybe even intriguing short-term blips. Why is there a dark patch for breakingPeace in the 1840s and for damage in the 1880s, for example?  





Future questions
-----------

As I've mentioned, a further question to be explored is the relationship between these prosecution patterns and London's population growth. Another is the relationship between prosecutions in the Old Bailey and other London courts. The [Digital Panopticon](https://www.digitalpanopticon.org), another project I've recently worked on, has digitised a series of records for the [Middlesex Sessions from 1836-1889](https://www.digitalpanopticon.org/Middlesex_House_of_Detention_Calendars_1836-1889), and I'll be able to take a look at those and do some comparisons.
 

Resources
---------


<!-- The dataset and R code for the post are available at the site's [Github repository](https://github.com/sharonhoward/ihme_data) [201804_obo_1]. -->

[About the Old Bailey Online project](https://www.oldbaileyonline.org/about/project)

* [XML files and documentation (CC BY-NC)](http://dx.doi.org/10.15131/shef.data.4775434)

Inspirations for the heatmaps:

* [A brief visual history of MARC cataloging at the Library of Congress](http://sappingattention.blogspot.com/2017/05/a-brief-visual-history-of-marc.html)
* [Time Based Heatmaps in R](https://www.littlemissdata.com/blog/heatmaps)




Technical notes
--------

My current workflow for producing tabular data from the OBO XML has two steps:

1. Query the XML using [BaseX](http://basex.org/) to extract data including offence category (or categories), defendant gender(s) and session date for every trial in the Proceedings and exported it as CSV. (Tools note: BaseX is a very useful cross-platform XML database app, which I heartily recommend. It's lightweight but can handle large amounts of data.)
2. Import the CSV data into a MySQL database for additional work: in this case, for each trial, I calculated offence and defendant counts; I also added trial word counts (which I'd made separately). For the sake of simplicity in this analysis, I've also added a column that contains only the *first* listed offence category for each trial, and did the same thing for defendant gender. (Tools note: my MySQL database app of choice on a Mac is SequelPro; I haven't yet found one I like as much for Windows but generally use HeidiSQL.)

For this post, I exported the data back out again to a CSV file to work with in RStudio. (There are R packages like dbplyr to query the database directly, but I want to keep this site as simple and self-contained as I can.)

In R itself I've done a bit of cleaning/tidying/filtering and I've added some new variables: a) year of trial, and a properly date-formatted session date; and b) a trial type based on the defendant count ('single' for one individual or 'group' for two or more). In the future I'll probably do more of step 2 directly in R, but in order to get going I wanted to use data I already had to hand that didn't need much additional work.
