---
title: "Pairing up with Observable Plot"
subtitle: "Different ways of visualising paired data"
author: "Sharon Howard"
date: "2025-08-09"
categories: [interactive, methods]
code-fold: true
execute: 
  warning: false
  message: false
---

## Introduction

I've started exploring [Quarto's built in support for Observable JS](https://quarto.org/docs/interactive/ojs/) to make interactive charts. Here I'm working through some options for visualising "paired" data (where there are two groups for comparison: eg male v female, two points in time, or some other category with just two values).


## data

- [Cheshire petitions 1573-1798](https://doi.org/10.5281/zenodo.7027692)
- [York defamation causes 1661-1700](https://sharonhoward.org/defamation/causes.html)
- [Registers of habitual criminals 1881-1925](https://figshare.com/articles/dataset/Metropolitan_Police_Register_of_Habitual_Criminals_1881-1925/5697994)
- [Monarchs and Marriages](https://www.ianvisits.co.uk/articles/a-list-of-monarchs-by-marriage-6857/) ([via](https://nrennie.rbind.io/scrollytelling/posts/monarchs/))
 
 
 
```{r}
library(tidyverse)
library(mindseyedata)

petitions_slope <-
cheshire_petitions |>
  filter(year %in% c(1628, 1668) & topic !="other") |>
  count(topic, year, name="petitions") |>
  filter(n()==2, .by = topic)

petitions_topic_gender <-
cheshire_petitions |>
  filter(petition_gender %in% c("f", "m") 
         & !str_detect(petition_type, "collective") 
         & !topic %in% c("other")) |>
  mutate(gender = if_else(petition_gender=="m", "male", "female"))

petitions_dumbel <- 
cheshire_petitions |>
  filter(topic != "other") |>
  mutate(petition_type = str_remove(petition_type, " *on behalf")) |>
  mutate(petition_type_s = if_else(petition_type=="single", "single", "multiple")) |>
  count(topic, petition_type_s) |>
  group_by(topic) |>
  mutate(min = min(n), max=max(n)) |>
  ungroup()

york_causes_csv <-
  read_csv(here::here("site_data/sh/york_causes_250809.csv") ) 

york_causes <-
york_causes_csv |>
  select(uid, year, pltf_gender, def_gender) |>
  pivot_longer(c(pltf_gender, def_gender), 
               names_to = "person", values_to = "gender") |>
  mutate(gender = if_else(gender=="m", "male", "female")) 



monarchs_marriages <-
read_tsv(here::here("site_data/other/monarchs-marriages.tsv")) |>
  janitor::clean_names() |>
  fill(monarch) |>
  select(-age_diff) |>
  mutate(m_gender = case_when(
    str_detect(monarch, "Mary|Elizabeth|Ann|Victoria") ~ "f",
    .default = "m"
  )) |>
  mutate(c_gender = if_else(m_gender=="f", "m", "f")) |>
  group_by(monarch) |>
  arrange(year_of_marriage, .by_group = TRUE) |>
  mutate(marriage_id = row_number()) |>
  ungroup() |>
  mutate(marriage_id = glue::glue("{m_gender}_{c_gender}_{year_of_marriage}_{marriage_id}")) |>
  mutate(monarch = glue::glue("{monarch} ({m_age})"), consort = glue::glue("{consort} ({c_age})")) |>
  relocate(marriage_id) |>
  mutate(f = case_when(
    m_gender=="f" ~ m_age,
    c_gender=="f" ~ c_age
  )) |>
  mutate(m = case_when(
    m_gender=="m" ~ m_age,
    c_gender=="m" ~ c_age
  )) 

set.seed(1357)
rhc_sampled <-
  bind_rows(
    rhc |>
      filter(gender=="male" & !is.na(height_inches)) |>
      slice_sample(prop = 0.15) ,
    rhc |>
      filter(gender=="female" & !is.na(height_inches))
  )
```



```{r}
# get data ready for OJS
ojs_define(ojs_petitions_slope = petitions_slope)
ojs_define(ojs_petitions_dumbel = petitions_dumbel)
ojs_define(ojs_petitions_gender = petitions_topic_gender)
ojs_define(ojs_york = york_causes)
ojs_define(ojs_rhc = rhc_sampled)
ojs_define(ojs_monarchs_marriages_wide = monarchs_marriages)
```



```{ojs}
ojsPetitionsSlope = transpose(ojs_petitions_slope)
ojsPetitionsDumbel = transpose(ojs_petitions_dumbel)
ojsPetitionsGender = transpose(ojs_petitions_gender)
ojsYork = transpose(ojs_york)
ojsRhc = transpose(ojs_rhc)
ojsMarriagesMonarchsWide = transpose(ojs_monarchs_marriages_wide)
```


## Diverging stacked dots

These show frequencies as dots (one per observation), so they tend to be most useful for showing relatively small amounts of data. 

This shows the gender of people in York defamation causes per year. (It's unusually balanced for a set of early modern legal records!)

- [Observable Plot gallery example](https://observablehq.com/@observablehq/plot-stacked-dots) 

```{ojs}

Plot.plot({
  aspectRatio: 1,
  subtitle: "York defamation causes and gender",
  x: {label: "year",
      tickFormat: d3.format('d')},
  y: {
    grid: true,
    label: "← female · male →",
    labelAnchor: "center",
    tickFormat: Math.abs
  },
  color: {legend: true},
  marks: [
    Plot.dot(
      ojsYork,
      Plot.stackY2({
        x: "year",
        y: (d) => d.gender === "male" ? 1 : -1,
        fill: "gender",
        r:4,
        //title: "full_name"
      })
    ),
    Plot.ruleY([0])
  ]
})

```



## Diverging stacked bars

These are often used for visualising [Likert scale](https://en.wikipedia.org/wiki/Likert_scale) responses to survey questions, as in the 
[Observable Plot gallery example](https://observablehq.com/@observablehq/plot-diverging-stacked-bar), or for [population pyramids](https://en.wikipedia.org/wiki/Population_pyramid). They can show much larger amounts of data than the stacked dots.

```{ojs}
Plot.plot({
  x: {label: "petitions",
      labelAnchor: "center",
      labelArrow: "none",
      tickFormat: Math.abs},
  y: { tickSize: 0 },
  subtitle: "Cheshire petitions: compare topics by gender",
  marginLeft: 150,
  color: {
    legend: true
  },
  marks: [
    Plot.ruleX([0]),
    Plot.barX(
      ojsPetitionsGender,
      Plot.groupY(
        { x: (d) => d.length * sign(d[0].gender) },
        {fill: "gender",
          y: "topic" }
      )
    )
  ]
})
```


```{ojs}
// simplified version of gallery example
sign = (label) => label.match(/female/i) ? -1 : 1
```



## Overlapping histograms 

Rather neat way to show a pair of histograms that overlap, like heights or weights for men and women. 

However, if using a simple count, it does need the numbers of observations to be roughly similar. In the rhc data there are a lot more men than women, so I've simply taken a 15% sample of the men to get more even numbers. If I were doing this properly, I'd need to look into better normalisation methods.

- [Observable Plot gallery example](https://observablehq.com/@observablehq/plot-overlapping-histogram)


```{ojs}
Plot.plot({
  subtitle: "Heights of male and female convicts",
  round: true,
  color: {legend: true},
  marks: [
    Plot.rectY(ojsRhc, Plot.binX({y2: "count"}, {x: "height_inches", fill: "gender", mixBlendMode: "multiply"})),
    Plot.ruleY([0])
  ]
})
```


## Slopegraph

I've recently posted on slopegraphs in ggplot; it's handy to have an Observable version too.

- [Observable Plot gallery example](https://observablehq.com/@observablehq/plot-slope-chart)

```{ojs}
Plot.plot({
  height: 400,
  width: 400,
  x: {axis: "top", type: "ordinal", tickFormat: "", inset: 90, label: null},
  y: {axis: null, inset: 20},
  marks: [
    Plot.line(ojsPetitionsSlope, {x: "year", y: "petitions", z: "topic"}),
    d3.groups(ojsPetitionsSlope, (d) => d.year === 1628)
      .map(([left, petitions]) =>
        Plot.text(petitions, occlusionY({
          x: "year",
          y: "petitions",
          text: left
            ? (d) => `${d.topic} ${d.petitions}`
            : (d) => `${d.petitions} ${d.topic}`,
          textAnchor: left ? "end" : "start",
          dx: left ? -3 : 3,
          radius: 5.5
        }))
      )
  ],
  subtitle: "Compare petition topics in 1628 and 1668"
})
```

```{ojs}
// I've just copied this from the example; no idea how it actually works at this point!
// OcclusionY adds an initializer that shifts nodes vertically with a tiny force simulation.
occlusionY = ({radius = 6.5, ...options} = {}) => Plot.initializer(options, (data, facets, { y: {value: Y}, text: {value: T} }, {y: sy}, dimensions, context) => {
  for (const index of facets) {
    const unique = new Set();
    const nodes = Array.from(index, (i) => ({
      fx: 0,
      y: sy(Y[i]),
      visible: unique.has(T[i]) // remove duplicate labels
        ? false
        : !!unique.add(T[i]),
      i
    }));
    d3.forceSimulation(nodes.filter((d) => d.visible))
      .force("y", d3.forceY(({y}) => y)) // gravitate towards the original y
      .force("collide", d3.forceCollide().radius(radius)) // collide
      .stop()
      .tick(20);
    for (const { y, node, i, visible } of nodes) Y[i] = !visible ? NaN : y;
  }
  return {data, facets, channels: {y: {value: Y}}};
})
```



## Dumbbell chart 

Another way to easily compare two sets of observations. It's quite similar to a slopegraph in some ways, a dotplot combined with a line to connect the pairs. It can be effective to reorder by size of the difference. 

- [Observable Plot gallery example](https://observablehq.com/@observablehq/plot-dot-plot)


```{ojs}
Plot.plot({
    subtitle: "Topics of petitions with single and multiple petitioners",
    marginLeft: 150,
    x: {grid: true, label:"petitions"},
    color: {legend: true, 
            range: ['#DDAA33', '#BB5566'], 
			      domain: ["single", "multiple"]
			      },
    marks: [
      Plot.ruleX([0]),
      Plot.ruleY(ojsPetitionsDumbel, {
        x1: "min", 
        x2: "max", 
        y: "topic",
        stroke: "lightGray",
        strokeWidth: 5,
      } ),
      Plot.dot(ojsPetitionsDumbel, {
        x: "n", 
        y: "topic", 
        fill: "petition_type_s", 
        r: 6
        })
    ]
  })
```


## Difference arrows 

[update: added 1/10/25]

- [Observable Plot gallery](https://observablehq.com/@observablehq/plot-difference-arrows) 

A fun plot; quite similar to a dumbbell, but with arrows to show direction of difference/change.

[Ian Mansfield](https://www.ianvisits.co.uk/articles/a-list-of-monarchs-by-marriage-6857/) put together a table of the ages of monarchs and their consorts at marriage (starting in the 9th century, but I've just pulled out the monarchs since 1066).

First, whether the monarch was the older or younger of the pair. You can hover over rows to see more details.

```{ojs}
Plot.plot({
  marginTop: 0,
  marginLeft: 100,
  y: {grid: true, label: null},
  
  color: {
    type: "categorical",
    domain: [-1, 1],
    range: ['#DDAA33', '#BB5566'],
    unknown: "#aaa",
    transform: Math.sign,
    tickFormat: (c) => c < 0 ? "monarch younger" : "monarch older", 
    legend: true
  },
  
  marks: [
    Plot.ruleX([0]),
    Plot.link(
      ojsMarriagesMonarchsWide,
        {
          x1: "m_age", // 
          x2: "c_age",
          stroke:  (d) => d.m_age - d.c_age, 
          y: "marriage_id",
          markerStart: "dot", 
          markerEnd: "arrow",
          channels: {year: 'year_of_marriage', "monarch": "monarch", "consort": "consort"}, 
          sort: {y: 'year'}, 
          strokeWidth: 2 ,
 
        // tooltip
  			tip: {
    			format: {
    				x: false, 
      			y: false, 
      			stroke: false,
    				"year": (d) => `${d}`,
    			},
  				anchor: "left"
  		  }        
      }
    )
  ]
})

```

Second, whether the wife was the younger or older.

```{ojs}
Plot.plot({
  marginTop: 0,
  marginLeft: 100,
  y: {grid: true, label: null},
  
  color: {
    type: "categorical",
    domain: [-1, 1],
    unknown: "#aaa",
    transform: Math.sign,
    tickFormat: (c) => c < 0 ? "wife older" : "wife younger", 
    legend: true
  },
  
  marks: [
    Plot.ruleX([0]),
    Plot.link(
      ojsMarriagesMonarchsWide,
        {
          x1: "f",
          x2: "m",
          stroke:  (d) => d.m - d.f, // negative if f is older.
          y: "marriage_id",
          markerStart: "dot", 
          markerEnd: "arrow", 
          channels: {year: 'year_of_marriage', monarch: "monarch", consort: "consort"}, 
          sort: {y: 'year'}, 
          strokeWidth: 2,
        // tooltip
  			tip: {
    			format: {
    				x: false, 
      			y: false, 
      			stroke: false,
    				"year": (d) => `${d}`,
    			},
  				anchor: "left"
  		  }     
      }
    )
  ]
})

```



