---
title: "Colour Notes"
author: "Sharon Howard"
date: "2025-03-13"
categories: [accessibility, colour, methods]
execute: 
  warning: false
  message: false
---

I haven't always been diligent in ensuring visualisation colours work for everyone, but recently working with a colourblind colleague forced me to pull my socks up. 

I plan to review past work on this blog and make improvements where I can; this post is a reference for me in the first instance, and I hope might be useful to others as well.

## swatchplot()

```{r}
library(colorspace)
```

I've probably been using the [colorspace](https://colorspace.r-forge.r-project.org/index.html) swatchplot() function more than anything else to check palettes, especially custom palettes that I'd use in ggplot's scale_*_manual functions. 

There are other tools I use (see lists below), but for me this is the quickest one for quickly testing a palette within R. (Some other packages have similar simulators which I haven't tried yet.)

It's handy with named palettes too; here's the [viridis palette](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) (specifying 8 colours), with simulation of all three of the main colour blindness types to facilitate comparison:

```{r}
swatchplot(sequential_hcl(8, "Viridis"), cvd = TRUE)
```

And for a diverging palette ([ColorBrewer](https://colorbrewer2.org/#type=diverging&scheme=PRGn&n=9)):


```{r}
swatchplot(diverging_hcl(7, "Purple-Green"), cvd = TRUE)
```

But I think swatchplot really comes into its own if you want to test and tweak a custom palette. You can just give it a vector of colours. 

This is a red-green-blue combination. It's... not good.

```{r}
swatchplot(c("#F8766D", "#00BA38", "#619CFF"), cvd=TRUE)
```

Something better (colours from Paul Tol's light qualitative scheme).

```{r}
swatchplot(c("#EE8866", "#EEDD88", "#99DDFF", "#44BB99", "#BBCC33"), cvd=TRUE)
```

If you need a three colour scheme, these are the colours in Paul Tol's high-contrast palette:

```{r}
swatchplot(c('#DDAA33', '#BB5566',  "#0072B2"), cvd=T)
```

## cvd_grid()

With [colorblindr](https://github.com/clauswilke/colorblindr) you can go a step further than the swatchplot and simulate actual visualisations; this may be particularly useful for line graphs and scatterplots where you can get rather different effects than the blocks of colour in bar charts and the like.

```{r}
# install colorblindr from github
#remotes::install_github("clauswilke/colorblindr")

library(ggthemes)
library(colorblindr)
library(ggplot2)
library(palmerpenguins)
```

Scatterplot of the Palmers Penguins dat, using the default ggplot colour palette (as a general rule: don't do this; it's ugly as well as being abysmal for the colourblind).

```{r}
pp <-
penguins |>
  dplyr::filter(!is.na(sex) ) |>
  ggplot(aes(bill_length_mm, bill_depth_mm, colour = species)) +
  geom_point()

pp
```

Simulations:

- R default colours

```{r}
cvd_grid(pp)
```



- the Okabe-Ito colorblind palette in ggthemes

```{r}
cvd_grid(pp + scale_colour_colorblind())
```

- the Paul Tol palette in ggthemes:

```{r}
cvd_grid(pp + scale_colour_ptol())
```


## Further resources and reading

Useful tools:

- [Obumbratta colour tool](https://obumbratta.com/colour)
- [Viz Palette](https://projects.susielu.com/viz-palette)
- [An interactive guide to color and contrast](https://colorandcontrast.com/#/)
- [Coblis color blindness simulator](https://www.color-blindness.com/coblis-color-blindness-simulator/)

Blog posts etc:

- Datawrapper blog has a great ongoing series of posts, including several specifically on color blindness
  - [Color in data vis category](https://www.datawrapper.de/blog/category/color-in-data-vis)
  - [Color blindness: Part 1](https://www.datawrapper.de/blog/colorblindness-part1)
  - [Color blindness: Part 2](https://www.datawrapper.de/blog/colorblindness-part2)
  - [Color blindness: Part 3](https://www.datawrapper.de/blog/colorblindness-part3)

- [Quickly assessing colour palettes](https://www.njtierney.com/post/2020/10/15/assess-colour/)
- [Crafting an effective data visualization color palette (Observable)](https://observablehq.com/blog/crafting-data-colors)
- [Coloring in R's blind spot](https://www.zeileis.org/news/coloring/)
- [Working with colours in R](https://nrennie.rbind.io/blog/colours-in-r/)

Some R packages:

- [prismatic](https://emilhvitfeldt.github.io/prismatic/) - "The goal of prismatic is to provide color manipulation tools in R, in an intuitive, low-dependency and functional way"
- [paleteer](https://emilhvitfeldt.github.io/paletteer/) - "The goal of paletteer is to be a comprehensive collection of color palettes in R using a common interface"
- [khroma](https://packages.tesselle.org/khroma/) - "an implementation of Okabe and Ito, Tol and Crameri color schemes"
- [Colors for all](https://cols4all.github.io/cols4all-R/) - "palettes from several popular and lesser known color palette series... scored on several aspects"


Okabe and Ito:

- [Color universal design: How to make figures and presentations that are friendly to Colorblind people](https://jfly.uni-koeln.de/color/)
- also: [the Okabe and Ito palette as hex values](https://mikemol.github.io/technique/colorblind/2018/02/11/color-safe-palette.html)

Paul Tol:

- [Paul Tol's notes](https://sronpersonalpages.nl/~pault/)

Beyond colour blindness, I've been reading Jonathan Godfrey's work for blind R users:

- [BrailleR](https://ajrgodfrey.github.io/BrailleR/) [R package]
- [Statistical software and blind users](https://r-resources.massey.ac.nz/StatSoftware/) [resource]